---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Author: Jennifer Campbell  
Class: Reproducible Research, week 2, project 1  
Date: May 20, 2017  
Aim: Write a report that answers the questions detailed below. Ultimately, complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.  
This analysis has been conducted on a Windows machine running on Windows 10.  

## Loading and preprocessing the data
In the first step of the analysis, check whether the input data file activitydata.csv is in the working directory, 
and then download and unzip the file if necessary:

``` {r, echo=TRUE}
if(!file.exists("activity.csv")){
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url, destfile='activity.zip', method="wininet", mode="wb")
    unzip(zipfile = "activity.zip")
}
```

Then we can read in the data:  
```{r echo=TRUE}
df <- read.csv("activity.csv", 
                na.strings = "NA" header = TRUE, colClasses = c("Numeric","Character","Numeric"))
```                

There are a total of 17,568 observations in this dataset. 
The variables included in this dataset are:

### Codebook  
| Field | Description|
|:------|:-----------|
|Steps | Number of steps taken in a 5-minute interval (i.e. 288 observations per day) <br> Numeric variable <br> Range: 0 to 806; 2,304 missing ("NA") values|
|Date | The date on which the measurement was taken in YYYY-MM-DD format <br> Character variable <br> Range: 2012-10-01 to 2012-11-30 (61 days); No missing values|
|Interval | Identifier for the 5-minute interval in which measurement was taken <br> Numeric variable <br> Range: 0 to 2,355; No missing values|

We can use the function `aggregate()` to collapse data in order to analyze the observations on a 'per day' basis, and simultaneously exclude missing values (this is the function default, but it is explicitly coded here).

```{r echo=TRUE}
dailydata <- aggregate(steps ~ date, data=df, FUN=sum, na.action = na.omit)
```

## What is mean total number of steps taken per day?
  
The distribution of the daily steps data can be represented with a histogram, which can be generated as follows:

```{r echo=TRUE}
hist(dailydata$steps, main="Histogram of Daily Steps", xlab="Total Daily Steps")
```

We can also summarize the data with the following measures:

```{r echo=TRUE}
summary(dailydata$steps)
dailymean <- mean(dailydata$steps)
dailymedian <- median(dailydata$steps)
```
 
This provides us with the mean number of steps per day of `r dailymean` and median number of steps per day of `r dailymedian`.  
 
## What is the average daily activity pattern?
 
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 
 
## Imputing missing values  

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps? 
 
 
## Are there differences in activity patterns between weekdays and weekends?  

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).


