---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Author: Jennifer Campbell  
Class: Reproducible Research, week 2, project 1  
Date: May 20, 2017  
Aim: Write a report that answers the questions detailed below. Ultimately, complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.  
This analysis has been conducted on a Windows machine running on Windows 10.  

## Loading and preprocessing the data
In the first step of the analysis, check whether the input data file activitydata.csv is in the working directory, 
and then download and unzip the file if necessary:

``` {r, echo=TRUE}
if(!file.exists("activity.csv")){
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url, destfile='activity.zip', method="wininet", mode="wb")
    unzip(zipfile = "activity.zip")
}
```

Then we can read in the data:  
```{r echo=TRUE}
df <- read.csv("activity.csv", 
                na.strings = "NA" header = TRUE, colClasses = c("Numeric","Character","Numeric"))
```                

There are a total of 17,568 observations in this dataset. 
The variables included in this dataset are:

### Codebook  
| Field | Description|
|:------|:-----------|
|Steps | Number of steps taken in a 5-minute interval (i.e. 288 observations per day) <br> Numeric variable <br> Range: 0 to 806; 2,304 missing ("NA") values|
|Date | The date on which the measurement was taken in YYYY-MM-DD format <br> Character variable <br> Range: 2012-10-01 to 2012-11-30 (61 days); No missing values|
|Interval | Identifier for the 5-minute interval in which measurement was taken <br> Numeric variable <br> Range: 0 to 2,355; No missing values|

We can use the function `aggregate()` to collapse data in order to analyze the observations on a 'per day' basis, and simultaneously exclude missing values (this is the function default, but it is explicitly coded here).

```{r echo=TRUE}
dailydata <- aggregate(steps ~ date, data=df, FUN=sum, na.action = na.omit)
```

## What is mean total number of steps taken per day?
  
The distribution of the daily steps data can be represented with a histogram, which can be generated as follows:

```{r echo=TRUE}
hist(dailydata$steps, main="Histogram of Daily Steps", xlab="Total Daily Steps")
```

We can also summarize the data with the following measures:

```{r echo=TRUE}
summary(dailydata$steps)
dailymean <- mean(dailydata$steps)
dailymedian <- median(dailydata$steps)
```
 
 This provides us with the mean of 'r dailymean' and median of 'r dailymedian'
## What is the average daily activity pattern?
 
 
 
## Imputing missing values
 
 
 
## Are there differences in activity patterns between weekdays and weekends?
