---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Author: Jennifer Campbell  
Class: Reproducible Research, week 2, project 1  
Date: May 20, 2017  
Aim: Write a report that answers the questions detailed below. Ultimately, complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.  
This analysis has been conducted on a Windows machine running on Windows 10

## Loading and preprocessing the data
In the first step of the analysis, check whether the input data file activitydata.csv is in the working directory, 
and then download and unzip the file if necessary:

``` {r, echo=TRUE}
if(!file.exists("activity.csv")){
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url, destfile='activity.zip', method="wininet", mode="wb")
    unzip(zipfile = "activity.zip")
}
```

Then we can read in the data:  
```{r echo=TRUE}
df <- read.csv("activity.csv", 
                na.strings = "NA" header = TRUE, colClasses = c("Numeric","Character","Numeric"))
```                

There are a total of 17,568 observations in this dataset. 
The variables included in this dataset are:

### Codebook  
| Field | Description|
|-------|------------|
|Steps | Number of steps taking in a 5-minute interval (missing values are coded as NA) |
|Date | The date on which the measurement was taken in YYYY-MM-DD format |
|Interval | Identifier for the 5-minute interval in which measurement was taken |




## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
